---
title: "DESeq Analysis"
output: html_notebook
---
https://github.com/mousepixels/sanbomics/blob/main/tutorial_complex_Heatmap.Rmd
Modified code from above source


```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DESeq2")
#may need to install command line access in terminal [xcode-select --install]
```

```{r}
library(DESeq2)

library(ggplot2)
```

```{bash}
ls

```


```{r}
Counts <- read.delim("hisat2_counts.csv", header = TRUE, row.names = 1, sep = ",")


Org <- c( "DRM1.1","DRM1.2", "DRM1.3", "DRM2.1","DRM2.2", "DRM2.3","DRM3.1","DRM3.2", "DRM3.3","DRM4.1","DRM4.2", "DRM4.3","DRM5.1","DRM5.2", "DRM5.3", "DRM6.1","DRM6.2", "DRM6.3","DRM7.1","DRM7.2", "DRM7.3","DRM8.1","DRM8.2", "DRM8.3")
Counts <- Counts[,Org]

AllGenes <- rownames(Counts)

GCounts <- Counts[,c(7,8,9,10,11,12)] #YPD gene diff
CCounts <- Counts[,c(7,8,9,19,20,21)] #REF condition diff
BCounts <- Counts[,c(19,20,21,22,23,24)] #BPS gene diff

GCounts <- GCounts[which(rowSums(GCounts) > 0),]
CCounts <- CCounts[which(rowSums(CCounts) > 0),]
BCounts <- BCounts[which(rowSums(BCounts) > 0),]

condition <- factor(c("A","A","A","B","B","B"))
Gcoldata <- data.frame(row.names = colnames(GCounts), condition)
Ccoldata <- data.frame(row.names = colnames(CCounts), condition)
Bcoldata <- data.frame(row.names = colnames(BCounts), condition)

Gdds <- DESeqDataSetFromMatrix(countData = GCounts, colData = Gcoldata, design = ~condition)
Cdds <- DESeqDataSetFromMatrix(countData = CCounts, colData = Ccoldata, design = ~condition)
Bdds <- DESeqDataSetFromMatrix(countData = BCounts, colData = Bcoldata, design = ~condition)


Gdds <- DESeq(Gdds)
Cdds <- DESeq(Cdds)
Bdds <- DESeq(Bdds)

Gvsdata <- vst(Gdds, blind=FALSE)
Cvsdata <- vst(Cdds, blind=FALSE)
Bvsdata <- vst(Bdds, blind=FALSE)

plotPCA(Gvsdata, intgroup = "condition")
plotPCA(Cvsdata, intgroup = "condition")
plotPCA(Bvsdata, intgroup = "condition")

plotDispEsts(Gdds)
plotDispEsts(Cdds)
plotDispEsts(Bdds)

Gres <- results(Gdds, contrast = c("condition", "A", "B"))
Cres <- results(Cdds, contrast = c("condition", "A", "B"))
Bres <- results(Bdds, contrast = c("condition", "A", "B"))

Gsigs <- na.omit(Gres)
Csigs <- na.omit(Cres)
Bsigs <- na.omit(Bres)

write.csv(Gsigs, file = "YPDAllGenes_Results.csv")
write.csv(Csigs, file = "RefTreatAllGenes_Results.csv")
write.csv(Bsigs, file = "BPSAllGenes_Results.csv")

Gsigs <- Gsigs[Gsigs$padj < 0.05,]
Csigs <- Csigs[Csigs$padj < 0.05,]
Bsigs <- Bsigs[Bsigs$padj < 0.05,]

write.csv(Gsigs, file = "YPDGenedeseq_results.csv")
write.csv(Csigs, file = "RefTreatmentdeseq_results.csv")
write.csv(Bsigs, file = "BPSGenedeseq_results.csv")
```


```{r}
Gdf <- as.data.frame(Gsigs)
Cdf <- as.data.frame(Csigs)
Bdf <- as.data.frame(Bsigs)


gene_map <- read.csv("CLUSGenesByOrthologs.csv", header = TRUE)

keys <- gene_map$Input
values <- gene_map$Merged
l <- list()
for (i in 1:length(keys)){
  l[keys[i]] <- values[i]
}

```





```{r}
#for non-mapped labels
Gno_values <- setdiff(rownames(Gdf), keys)
for (i in 1:length(Gno_values)){
  l[Gno_values[i]] <- 'NA'
}
Cno_values <- setdiff(rownames(Cdf), keys)
for (i in 1:length(Cno_values)){
  l[Cno_values[i]] <- 'NA'
}
Bno_values <- setdiff(rownames(Bdf), keys)
for (i in 1:length(Bno_values)){
  l[Bno_values[i]] <- 'NA'
}


```


```{r}
Gdf$symbol <- unlist(l[rownames(Gdf)], use.names = FALSE)
Cdf$symbol <- unlist(l[rownames(Cdf)], use.names = FALSE)
Bdf$symbol <- unlist(l[rownames(Bdf)], use.names = FALSE)
 
```



```{r}
Gdf.top <- Gdf[ (Gdf$baseMean > 50) & (abs(Gdf$log2FoldChange) > 1),]
Cdf.top <- Cdf[ (Cdf$baseMean > 50) & (abs(Cdf$log2FoldChange) > 1),]
Bdf.top <- Bdf[ (Bdf$baseMean > 50) & (abs(Bdf$log2FoldChange) > 1),]

#order by logfoldchange
Gdf.top <- Gdf.top[order(Gdf.top$log2FoldChange, decreasing = TRUE),]
Cdf.top <- Cdf.top[order(Cdf.top$log2FoldChange, decreasing = TRUE),]
Bdf.top <- Bdf.top[order(Bdf.top$log2FoldChange, decreasing = TRUE),]


write.csv(Gdf.top, file = "GTopRes.csv")
write.csv(Cdf.top, file = "CtopRes.csv")
write.csv(Bdf.top, file = "BtopRes.csv")

GdfTopRes <- read.csv("GTopResNAcorrect.csv", header = TRUE)
CdfTopRes <- read.csv("CtopRes.csv", header = TRUE)
BdfTopRes <- read.csv("BtopResNAcorrect.csv", header = TRUE)



```



```{r}
Grlog_out <- rlog(Gdds, blind=FALSE) #get normalized count data from dds object
Gmat<-assay(Grlog_out)[rownames(Gdf.top), rownames(Gcoldata)] #sig genes x samples
colnames(Gmat) <- rownames(Gcoldata)
Gbase_mean <- rowMeans(Gmat)
Gmat.scaled <- t(apply(Gmat, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(Gmat.scaled)<-colnames(Gmat)

Crlog_out <- rlog(Cdds, blind=FALSE) #get normalized count data from dds object
Cmat<-assay(Crlog_out)[rownames(Cdf.top), rownames(Ccoldata)] #sig genes x samples
colnames(Cmat) <- rownames(Ccoldata)
Cbase_mean <- rowMeans(Cmat)
Cmat.scaled <- t(apply(Cmat, 1, scale)) #center and scale each column (Z-score) then transpose
colnames(Cmat.scaled)<-colnames(Cmat)

Brlog_out <- rlog(Bdds, blind=FALSE)
Bmat<-assay(Brlog_out)[rownames(Bdf.top), rownames(Bcoldata)]
colnames(Bmat) <- rownames(Bcoldata)
Bbase_mean <- rowMeans(Bmat)
Bmat.scaled <- t(apply(Bmat, 1, scale))
colnames(Bmat.scaled)<-colnames(Bmat)

```


```{r}
Gnum_keep <- 44
Cnum_keep <- 11
Bnum_keep <- 36

Grows_keep <- c(seq(1:Gnum_keep))
Crows_keep <- c(seq(1:Cnum_keep))
Brows_keep <- c(seq(1:Bnum_keep))
```



```{r}
Gl2_val <- as.matrix(Gdf.top[Grows_keep,]$log2FoldChange) #getting log2 value for each gene we are keeping
colnames(Gl2_val)<-"logFC"
#Gmean <- as.matrix(Gdf.top[Grows_keep,]$baseMean) #getting mean value for each gene we are keeping
colnames(Gmean)<-"AveExpr"

Cl2_val <- as.matrix(Cdf.top[Crows_keep,]$log2FoldChange) #getting log2 value for each gene we are keeping
colnames(Cl2_val)<-"logFC"
#Cmean <- as.matrix(Cdf.top[Crows_keep,]$baseMean) #getting mean value for each gene we are keeping
colnames(Cmean)<-"AveExpr"

Bl2_val <- as.matrix(Bdf.top[Brows_keep,]$log2FoldChange) #getting log2 value for each gene we are keeping
colnames(Bl2_val)<-"logFC"
#Bmean <- as.matrix(Bdf.top[Brows_keep,]$baseMean) #getting mean value for each gene we are keeping
colnames(Bmean)<-"AveExpr"
```





```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
```


```{r}
#maps values between b/w/r for min and max l2 values
Gcol_logFC <- colorRamp2(c(min(Gl2_val),0, max(Gl2_val)), c("blue", "white", "red")) 
Ccol_logFC <- colorRamp2(c(min(Cl2_val),0, max(Cl2_val)), c("blue", "white", "red"))
Bcol_logFC <- colorRamp2(c(min(Bl2_val),0, max(Bl2_val)), c("blue", "white", "red")) 
#maps between 0% quantile, and 75% quantile of mean values --- 0, 25, 50, 75, 100

#Gcol_AveExpr <- colorRamp2(c(quantile(Gmean)[1], quantile(Gmean)[4]), c("white", "red"))

#Ccol_AveExpr <- colorRamp2(c(quantile(Cmean)[1], quantile(Cmean)[4]), c("white", "red"))

#Bcol_AveExpr <- colorRamp2(c(quantile(Bmean)[1], quantile(Bmean)[4]), c("white", "red"))

```



```{r}
ha <- HeatmapAnnotation(summary = anno_summary(gp = gpar(fill = 2), 
                                               height = unit(1, "cm")))

Gh1 <- Heatmap(Gmat.scaled[Grows_keep,], cluster_rows = F, ,
            column_labels = colnames(Gmat.scaled), name="Z-score", 
            cluster_columns = T, width = unit(2, "cm"), height = unit(15,"cm"))
Gh2 <- Heatmap(Gl2_val, row_labels = GdfTopRes$symbol[Grows_keep], 
            cluster_rows = F, 
            name="logFC",
            top_annotation = ha, 
            col = Gcol_logFC,
            cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
              grid.text(round(Gl2_val[i, j],2), x, y, gp=gpar(fontsize=8, col="black"))
            }, width = unit(1, "cm"))
#Gh3 <- Heatmap(Gmean, row_labels = Gdf.top$symbol[Grows_keep], 
        #    cluster_rows = F, name = "AveExpr", col=Gcol_AveExpr, width = unit(1, "cm"),
        #    cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
         #     grid.text(round(Gmean[i, j],2), x, y, gp=gpar(fontsize=7, col="black"))
        #    })
Gh<-Gh1 + Gh2
Gh


#---------------------------------------------


Ch1 <- Heatmap(Cmat.scaled[Crows_keep,], cluster_rows = F, 
            column_labels = colnames(Cmat.scaled), name="Z-score",
            cluster_columns = T, width = unit(2, "cm"), height = unit(5,"cm"))
Ch2 <- Heatmap(Cl2_val, row_labels = CdfTopRes$symbol[Crows_keep], 
            cluster_rows = F, name="logFC", top_annotation = ha, col = Ccol_logFC,
            cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
              grid.text(round(Cl2_val[i, j],2), x, y, gp=gpar(fontsize=8, col="black"))
            }, width = unit(1, "cm"))
#Ch3 <- Heatmap(Cmean, row_labels = Cdf.top$symbol[Crows_keep], 
         #   cluster_rows = F, name = "AveExpr", col=Ccol_AveExpr, width = unit(1, "cm"),
          #  cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
           #  grid.text(round(Cmean[i, j],2), x, y, gp=gpar(fontsize=7, col="black"))
           # })
Ch<-Ch1+Ch2
Ch

#-------------------------------------
Bh1 <- Heatmap(Bmat.scaled[Brows_keep,], cluster_rows = F, 
            column_labels = colnames(Bmat.scaled), name="Z-score",
            cluster_columns = T, width = unit(2, "cm"), height = unit(12,"cm"))
Bh2 <- Heatmap(Bl2_val, row_labels = BdfTopRes$symbol[Brows_keep], 
            cluster_rows = F, name="logFC", top_annotation = ha, col = Bcol_logFC,
            cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
              grid.text(round(Bl2_val[i, j],2), x, y, gp=gpar(fontsize=8, col="black"))
            }, width = unit(1, "cm"))
#Bh3 <- Heatmap(Bmean, row_labels = Bdf.top$symbol[Brows_keep], 
          #  cluster_rows = F, name = "AveExpr", col=Bcol_AveExpr, width = unit(1, "cm"),
          #  cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
           #   grid.text(round(Bmean[i, j],2), x, y, gp=gpar(fontsize=7, col="black"))
          #  })
Bh<-Bh1+Bh2
Bh
```


```{r}
png("./heatmap_vGene.png", res = 300, width = 3000, height = 5500)
print(Gh)
dev.off()

png("./heatmap_vTreatment.png", res = 300, width = 3000, height = 5500)
print(Ch)
dev.off()

png("./heatmap_vBPS.png", res = 300, width = 3000, height = 5500)
print(Bh)
dev.off()

```


```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

